% Chapter 1

\chapter{Sources} % Main chapter title
[1] CDC https://www.cdc.gov/lyme/stats/index.html \newline

\noindent [2] Record, Nick \newline

\noindent [3] Elith, Jane Leathwick, John R"Species Distribution Models: Ecological Explanation and Prediction Across Space and Time" https://doi.org/10.1146/annurev.ecolsys.110308.120159\newline

\noindent [4] Manuel Gimond, "Introduction to GIS", Colby College \newline

\noindent [5] J. S. Gray, H. Dautel, A. Estrada-Peña, O. Kahl, and E. Lindgren, “Effects of Climate Change on Ticks and Tick-Borne Diseases in Europe,” Interdisciplinary Perspectives on Infectious Diseases, vol. 2009, Article ID 593232, 12 pages, 2009. doi:10.1155/2009/593232\newline

\noindent [6] A statistical explanation of MaxEnt for ecologists Jane Elith, Steven J. Phillips, Trevor Hastie, Miroslav Dudi, Yung En Chee and Colin J. Yates\newline

\noindent [7] A. Townsend Peterson, Monica Papes, Jorge Sobero, "Rethinking receiver operating characteristic analysis applications in ecological niche modeling" \newline

\noindent [8] Peterson, A.T., Papes, M., Eaton, M., 2007. Transferability and model evaluation in ecological niche modeling: a comparison of GARP and Maxent. Ecography 30, 550–560. \newline

\noindent [9] OMRI ALLOUCHE, ASAF TSOAR and RONEN KADMON. Assessing the accuracy of species distribution models: prevalence, kappa and the true skill statistic (TSS) \newline

\noindent [10]  Axiomatic Derivation of the Principle of Maximum Entropy and the Principle of Minimum Cross-Entropy JOHN E. SHORE, MEMBER, IEEE, AND RODNEY w. JOHNSON \newline

\noindent [11] Berger, Adam. \newline http://www.cs.cmu.edu/afs/cs/user/aberger/www/html/tutorial/node2.htmlSECTION00011000000000000000 \newline

\noindent [12] S.J. Phillips et al. Maximum entropy modeling of species geographic distributions \newline

\noindent[13] Spatial Point Patterns Methodology and Applications with R
Adrian Baddeley-Ege Rubak-Rolf Turner - Crc Press - 2016 \newline

\noindent [14] https://mgimond.github.io/ES218/Week01.html